---
title: 'Analysis of Recruited monocytes and macrophages (RecM)'
author:
- name: Anson Wong
  affiliation: Molecular Immunity, Murdoch Children's Research Institute
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(phateR)
library(slingshot)
library(readr)
library(cowplot)
library(scattermore)
library(ggrastr)
library(viridis)
library(scater)
library(circlize)
library(ggplotify)
library(tibble)
library(ComplexHeatmap)
library(RColorBrewer)
})
set.seed(1990)

```

# 2 Annotation, integration, clustering, finding marker genes
Cells annotated as RecM in the combined dataset were extracted and re-clustered.
Methods for normalization, integration, clustering and annotation were same as before, except that a resolution of 2 was selected.

For details, refer to code/bal_RecM.integration_clustering.R

# 3 Manual annotation - RecM
## 3.1 Load clustered object
```{r}
seu <- qread(here("data",
                  "SCEs",
                  "clustering",
                  "bal_RecM.integrated.clustered.SEU.qs"),nthreads=16)

seu$predicted.ann_finest_level <- factor(seu$predicted.ann_finest_level)
seu$sampleID <- factor(seu$sampleID)
seu$Capture <- factor(seu$Capture)
seu$Condition <- factor(seu$Condition)
seu$Bronchiectasis <- factor(seu$Bronchiectasis)

# SCT_snn_res.2 was selected for subclustering
levels(seu$SCT_snn_res.2)
seu$SCT_snn_res.3 <- factor(seu$SCT_snn_res.2,
                            levels=as.character(seq(1,27)))

# flip the UMAP x axis for easier visualization
seu@reductions$umap@cell.embeddings[,1] <- -seu@reductions$umap@cell.embeddings[,1]

```

## 3.2 Visualize UMAP embeddings
```{r}
# prepare colours
colours <- c(brewer.paired(12), brewer.pastel1(9), brewer.accent(8), brewer.set1(9),rev(brewer.dark2(8)), 
             brewer.set3(12))

p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "predicted.ann_finest_level") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p2 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        #label.box = TRUE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "SCT_snn_res.2") + #NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p3 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "celltype") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p4 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = TRUE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "SCT_snn_res.2") + NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p1
p2
p3
p4

```

## 3.3 Visualize proportion of cells per cluster
```{r}

labels <- c("predicted.ann_finest_level","SCT_snn_res.2")
p <- vector("list",length(labels))

for(label in labels){
  seu@meta.data %>%
    ggplot(aes(x = !!sym(label), fill = !!sym(label))) +
    geom_bar() +
    geom_text(aes(label = ..count..), stat = "count",
              angle=90, colour = "black", size = 2) +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    NoLegend() + scale_fill_manual(values = rev(colours)) +
    labs(y = "No. Cells (log scale)") -> p1
  
  seu@meta.data %>%
    dplyr::select(!!sym(label), Condition) %>%
    group_by(!!sym(label), Condition) %>%
    summarise(num = n()) %>%
    mutate(prop = num / sum(num)) %>%
  ggplot(aes(x = !!sym(label), y = prop * 100, 
             fill = Condition)) + 
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, size=6,
                                     vjust = 1, 
                                     hjust = 1)) +
    labs(y = "% Cells", fill = "Condition") -> p2
  
  p1 / p2 -> p[[label]]
}
p

```

## 3.4 Manual annotation
```{r}
subtype <- setNames(c("IM","MDM-PLA2G7high","MDM-FABP4high","MDM-PLA2G7high","MDM-FABP4high", # 1-5
                      "MDM-FABP4low","MDM-CCR2+","MDM-FABP4low","MDM-FABP4high","MDM-PLA2G7high", # 6-10
                      "MDM-FABP4low","MDM-FABP4low","MDM-PLA2G7high","MDM-FABP4high","MDM-IV", # 11-15
                      "MDM-FABP4low","MDM-FABP4high","MDM-FABP4high","IM","MDM-FABP4low", # 16-20
                      "Mono","MDM-FABP4low","Mono","MDM-FABP4low","MDM-FABP4high", # 21-25
                      "Mono","IM" # 26-27
                       ), levels(seu$SCT_snn_res.2))

subtype <- factor(setNames(subtype[seu$SCT_snn_res.2],colnames(seu)),
                     levels=c("Mono","MDM-IV","MDM-CCR2+","MDM-FABP4low","MDM-FABP4high","MDM-PLA2G7high","IM"
                              ))

seu <- AddMetaData(seu, metadata = subtype, col.name = "subtype")

# subcluster
subcluster <- setNames(as.character(1:7),levels(seu$subtype))

subcluster <- factor(setNames(subcluster[seu$subtype], colnames(seu)),
                     levels=as.character(1:7))

seu <- AddMetaData(seu, metadata = subcluster, col.name = "subcluster")

```

## 3.5 Save object
```{r}
celltype <- "RecM"
date <- Sys.Date()
if(!dir.exists(here("data","SCEs","final.annot"))) {
  dir.create(here("data","SCEs","final.annot"), recursive = TRUE)
}

out <- here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs"))
if(!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

```




# 4 Analysis
## 4.1 Read processed object
```{r}
# load analysis object
celltype <- "RecM"
seu <- qread(here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs")),
             nthreads=24)

```

## 4.2 Prepare object for downstream analysis
```{r}
# Subset object
seu <- subset(seu, Condition != "CF.ETI" & 
              sampleID != "M1C170C" & sampleID != "M1C176" & sampleID != "M1C180")

seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Add Condition2
draft1 <- factor(
      case_when(
        seu$Condition == "Control" ~ "HC"))
draft2 <- factor(
  case_when(
    seu$Bronchiectasis == "No" ~ "CFN",
    seu$Bronchiectasis == "Yes" ~ "CFB"
  )
)

Condition2 <- coalesce(draft1,draft2)
names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("HC","CFN","CFB"))

# Rename Condition
seu$Condition <- factor(
  case_when(
    seu$Condition == "Control" ~ "HC",
    seu$Condition == "CF" ~ "CF"
  ),
  levels=c("HC","CF")
)
# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", # healthy
                                "M1C170D","M1C176C","M1C180D","M1C177","M1C191B","M1C199B","M1C201A","M1C207B", # CFN
                                "M1C160(1)","M1C188","M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A" # CFB
                                ))

ID <- setNames(c(paste0("HC-0",1:6),paste0("CF-N",1:8),paste0("CF-B",1:8)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],levels=c(paste0("HC-0",1:6),paste0("CF-N",1:8),paste0("CF-B",1:8))),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Set colours
subcluster.colours <- setNames(c("#ef476f", "#7209b7", "#ffd166", "#83c5be",
                                 "#3a6ea5", "#9c6644", "#9e2a2b"),
                               levels(seu$subcluster))
subtype.colours <- setNames(c("#ef476f", "#7209b7", "#ffd166", "#83c5be",
                                 "#3a6ea5", "#9c6644", "#9e2a2b"),
                               levels(seu$subtype))
Condition2.colours <- setNames(c("#26547c","#ffd166","#ef476f"),
                               levels(seu$Condition2))
Condition.colours <- setNames(c("#26547c","#fb8b24"),
                              levels(seu$Condition))

subtype.labels <- c("Mono","MDM-IV",expression('MDM-'*italic('CCR2')^'+'),
                    expression('MDM-'*italic('PLA2G7')^'high'),
                    expression('MDM-'*italic('FABP4')^'low'),expression('MDM-'*italic('FABP4')^'high'),
                    "IM")
```

## 4.3 UMAP embeddings
```{r}
# grouped by subtype
p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        #cols=subtype.colours,
        #split.by = "Condition",
        group.by = "subtype") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
  scale_color_manual(values = subtype.colours,
                    labels = subtype.labels)

# grouped by severity
p2 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=Condition2.colours,
        group.by = "Condition2") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

# split by condition
p3 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subcluster.colours,
        split.by = "Condition",
        group.by = "subcluster") + NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

# split by severity
p4 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.2,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        split.by = "Condition2",
        group.by = "subtype") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text.align = 0,
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=14)) +
  scale_color_manual(values = subtype.colours,
                    labels = subtype.labels)

suppressWarnings(print(p4))

```

## 4.4 Proportion of cells
```{r}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$subtype,
                             sample=seu$ID)
                             
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("HC",42),rep("CF",112)),
                Condition2 = c(rep("HC",42),rep("CFN",56),rep("CFB",56))) -> df

df$Condition <- factor(df$Condition, levels=c("HC","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("HC","CFN","CFB"))

# stacked bar plot of cell proportion per sampleID
ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion of cells in RecM", fill="Subtype", x="Participant") +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = subtype.colours, labels = subtype.labels) +
  theme(legend.text.align = 0, legend.title = element_blank())

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(Freq ~ Condition2, 
              ref.group = "HC", 
              p.adjust.method = "BH")
stat.test
stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
p5 <- ggboxplot(df, x="clusters",
          y="Freq",
          fill="Condition2",
          palette = Condition2.colours#, outlier.shape=NA
          ) +
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01,
                     bracket.nudge.y = 0, vjust = 0, hide.ns = TRUE
                     ) +
  labs(y="Proportion of cells in RecM") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.title.x=element_blank(),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 10),
        legend.position = "right",
        legend.title = element_blank()) +
  scale_x_discrete(labels=subtype.labels)
p5
```


## 4.4 Marker gene analysis
### 4.4.1 Call marker genes with Cepo
```{r}
# Cepo
cepoMarkers <- Cepo(seu[["SCT"]]@data,
                    seu$subtype,
                    exprsPct = 0.2)
cepoMarkers$stats
cepoMarkers$stats <- cepoMarkers$stats[, c("Mono","MDM.IV","MDM.CCR2.","MDM.FABP4low","MDM.FABP4high","MDM.PLA2G7high","IM")]
cepoMarkers$stats <- cepoMarkers$stats[!grepl('^A[C|F|P|L][0|1]', row.names(cepoMarkers$stats)),]

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:200]
}) %>% data.frame() -> dat2

colnames(dat2) <- colnames(cepoMarkers$stats)
dat2

colnames(dat2) <- colnames(cepoMarkers$stats)
dat2 %>% data.frame() %>% write.csv(file=here("RecM.markers.csv"))
```

### 4.4.2 DotPlot of selected features
```{r fig.width=12,fig.height=6}
# Selected ---------------------------------------------------------------------
selected <- c("CD14","FCGR3A","FCN1","IL1B","S100A12",
              "S100A8","EREG","NAMPT","CD93","SERPINB9",
              "MARCKS","F13A1","MAF","FOLR2","RNASE1",
              "STAB1","PDGFC","MEF2C","MAFB","ITGA9",
              "SLC40A1","CCR2","CLEC5A","ITGAM","NLRP1",
              "NLRP12","VCAN","C1QA","C1QB","FABP4",
              "MARCO","CD163","MRC1","APOC1","PLA2G7",
              "LGMN","CD109","SPP1","MMP9","CD36",
              "MMP19","TGFBI","SDC3","GBP4","GBP5",
              "IDO1","CCL2","CCL8","CXCL9","CXCL10",
              "CXCL11","ISG15","ISG20","GCH1")

# function colors
alarmin <- "#E85D04"
cytokine <- "#D00000"
iron <- "#972D07"
inflammasome <- "#5A189A"
lipid <- "#003F88"
scavenger <- "#197278"
ecm <- "#FF5D8F"
others <- "#000000"

gene.colours <- setNames(c(others, others, ecm, cytokine,alarmin,
                           alarmin,rep(others,4),
                           others,rep(iron,3),others,
                           rep(iron,4),ecm,
                           iron, cytokine,others, others, inflammasome,
                           inflammasome,ecm,others,others,lipid,
                           rep(scavenger,3),lipid,ecm,
                           rep(ecm,5),
                           rep(ecm,3),rep(inflammasome,2),
                           others, rep(cytokine,4),
                           cytokine,rep(others,3)),
                         selected)
                           
# plot
# version 1: horizontal
p6 <- DotPlot(seu,
        features = selected,
        group.by = "subtype",
        col.max = 2,
        col.min = -2,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Arial"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   color=gene.colours,
                                   vjust = 1
                                   ),
        axis.text.y = element_text(size=10),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.ylorrd(9)[1], high=brewer.ylorrd(9)[8]) +
  scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p6$guides$colour$title <- "Mean Expression"
p6$guides$colour$title.position <- "top"
p6$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p6))

```


## 4.5 Trajectory analysis
### 4.5.1 Slingshot
```{r}
sce <- as.SingleCellExperiment(seu)

# Run Slingshot
set.seed(1990)
sce <- slingshot(sce,
                 clusterLabels = seu$subtype,
                 start.clus="Mono", #  ** define the starting cluster here **
                 stretch=0,
                 reducedDim = "UMAP",
                 allow.breaks=FALSE)

# prepare sds object
sds <- as.SlingshotDataSet(sce)

# add pseudotime results to object
slingPseudotime_1 <- setNames(sce$slingPseudotime_1, colnames(counts(sce)))
seu <- AddMetaData(seu, slingPseudotime_1, col.name = "slingPseudotime_1")

slingPseudotime_2 <- setNames(sce$slingPseudotime_2, colnames(counts(sce)))
seu <- AddMetaData(seu, slingPseudotime_2, col.name = "slingPseudotime_2")

slingPseudotime_3 <- setNames(sce$slingPseudotime_3, colnames(counts(sce)))
seu <- AddMetaData(seu, slingPseudotime_3, col.name = "slingPseudotime_3")

```

### 4.5.2 UMAP
```{r fig.width=9, fig.height=3}
embedded <- embedCurves(sce, "UMAP")

# pdt1
embedded1 <- slingCurves(embedded)[[1]]
embedded1 <- data.frame(embedded1$s[embedded1$ord,])
pdt1 <- (FeaturePlot(seu,
            pt.size = 0.001,
            #order = TRUE,
            features = 'slingPseudotime_1') + #NoLegend() +
  ggtitle("Lineage 1") +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(text=element_text(family="Arial"),
        axis.text = element_blank(),
        legend.text = element_blank(),
        legend.text.align = 0) +
  scale_color_gradient2(high = "#5ac2b1",low="#f7fcfb",na.value = "#d3d3d3",
                        guide = guide_legend(override.aes = list(alpha = 0)))) +
  geom_path(data=embedded1, aes(x=UMAP_1, y=UMAP_2), size=1.2)


# pdt2
embedded2 <- slingCurves(embedded)[[2]]
embedded2 <- data.frame(embedded2$s[embedded2$ord,])

pdt2 <- (FeaturePlot(seu,
            pt.size = 0.001,
            features = 'slingPseudotime_2') + #NoLegend() +
  ggtitle("Lineage 2") +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(text=element_text(family="Arial"),
        axis.text = element_blank(),
        legend.text = element_blank(),
        legend.text.align = 0) +
  scale_color_gradient2(high = "#456990",low="#f0f1f5",na.value = "#d3d3d3",
                        guide = guide_legend(override.aes = list(alpha = 0)))) +
  geom_path(data=embedded2, aes(x=UMAP_1, y=UMAP_2), size=1.2)

# pdt3
embedded3 <- slingCurves(embedded)[[3]]
embedded3 <- data.frame(embedded3$s[embedded3$ord,])

pdt3 <- (FeaturePlot(seu,
            pt.size = 0.001,
            features = 'slingPseudotime_3') + #NoLegend() +
  ggtitle("Lineage 3") +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(text=element_text(family="Arial"),
        axis.text = element_blank(),
        legend.text = element_blank(),
        legend.text.align = 0) +
  scale_color_gradient2(high = "#ee767a",low="#fff7f6",na.value = "#d3d3d3",
                        guide = guide_legend(override.aes = list(alpha = 0)))) +
  geom_path(data=embedded3, aes(x=UMAP_1, y=UMAP_2), size=1.2)

p7 <- pdt1|pdt2|pdt3
p7
```

### 4.5.3 Heatmap of marker genes along lineage 1
```{r fig.width=4, fig.height=4}
# Some code derived and adjusted from here: https://github.com/crickbabs/ZebrafishDevelopingHindbrainAtlas
createHeatmap <- function(obj, pseudotimeColName, title = "Trajectory", features = features) {
  # features
  Genedata = obj@assays$SCT@data
 # pseudotime
  pseudotimeData = obj@meta.data[pseudotimeColName] 
  colnames(pseudotimeData) = "pseudotime"
  pseudotimeData = pseudotimeData %>% 
    na.omit() %>% 
    rownames_to_column("cellID") %>% 
    arrange(pseudotime)
  
  # scale pseudotime values to 0-1
  scale <- function(x){(x-min(x))/(max(x)-min(x))}
  pseudotimeData$pseudotime <- scale(pseudotimeData$pseudotime)
  
  pt.matrix <- Genedata[features, pseudotimeData$cellID]
  pt.matrix <- t(apply(pt.matrix,1,function(x){smooth.spline(x,df=3)$y}))
  pt.matrix <- t(apply(pt.matrix,1,function(x){(x-mean(x))/sd(x)}))
  
  # rownames(pt.matrix) <- genes
  col_fun = colorRamp2(seq(from=0,to=1,length=2),c("#f7fcfb","#5ac2b1"))
  top_ha = HeatmapAnnotation(pseudotime=pseudotimeData$pseudotime,
                             col = list(pseudotime=col_fun),
                             show_legend = c(FALSE),
                             show_annotation_name = FALSE)
  bottom_ha = HeatmapAnnotation(subtype=obj@meta.data[pseudotimeData$cellID,"subtype"],
                                col = list(subtype=subtype.colours),
                                show_legend = c(FALSE),
                                show_annotation_name = FALSE)
  #column1_ha = HeatmapAnnotation(pseudotime=pseudotimeData$pseudotime)
  
  ht <- Heatmap(
    pt.matrix, 
    name                         = "z-score",
    col                          = colorRamp2(seq(from=-2,to=2,length=11),rev(brewer.pal(11, "RdYlBu"))),
    show_row_names               = TRUE,
    show_column_names            = FALSE,
    row_names_gp                 = gpar(fontsize = 8,fontface="italic"),
    # km = 6, if cluster_rows = TRUE
    row_title_rot                = 0,
    cluster_rows                 = FALSE, # TRUE 
    cluster_row_slices           = FALSE,
    cluster_columns              = FALSE,
    top_annotation               = top_ha,
    bottom_annotation            = bottom_ha)
  
  print(ht)
}

features <- c("HSPA1A","IL2RA","TRAF1","STAT4","S100A12",
              "SERPINB9","FCN1","RNASE1","CCR2","CLEC5A",
              "VCAN","ITGAM","APOC1","ACKR3","HP",
              "C1QB","FABP5","FABP4","LSAMP")

p8 <- createHeatmap(obj = seu, pseudotimeColName = "slingPseudotime_1", 
                    title = "Lineage 1 pseudotime", features=features)
p8
```

### 4.5.4 Density plot of cells along each lineage
```{r fig.width=8, fig.height=5}
subtype.list <- sds@lineages
names(subtype.list) <- c("slingPseudotime_1","slingPseudotime_2","slingPseudotime_3")

scale <- function(x){(x-min(x))/(max(x)-min(x))}
df.list <- lapply(names(subtype.list), FUN= function(x) {
  df <- seu@meta.data[c(x,"subtype","Condition2")]
  colnames(df) <- c("pseudotime","subtype","Condition2")
  df <- df %>% 
    na.exclude() %>%
    dplyr::filter(subtype %in% subtype.list[[x]])
  df$pseudotime <- scale(df$pseudotime)
  return(df)
})

names(df.list) <- c("Lineage 1","Lineage 2","Lineage 3")
df <- bind_rows(df.list, .id = "Lineage")
#df$pseudotime <- scale(df$pseudotime)

# plot density, grouped by subtype
pd1 <- ggplot(df, aes(x=pseudotime, y=..scaled.., fill=subtype, color=subtype)) +
  geom_density(alpha = 0.5, bw=0.1, adjust=0.6) +
  #scale_x_discrete(breaks=seq(0,1,0.25),labels=c("0","0.25","0.5","0.75","1")) +
  scale_fill_manual(values = subtype.colours, labels=subtype.labels) +
  scale_color_manual(values = subtype.colours, labels=subtype.labels) +
  theme_bw() +
  theme(strip.text.x = element_text(size=10),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        legend.text.align = 0,
        legend.position = 'right', 
        legend.box.just = "left"
        ) +
  facet_wrap(~Lineage) +
  guides(fill=guide_legend(title="RecM subtype"),
         color=guide_legend(title="RecM subtype")
         )

# plot density, grouped by Condition
pd2 <- ggplot(df, aes(x=pseudotime, y=..scaled.., color=Condition2)) +
  geom_density(alpha = 0.5, bw=0.1, adjust=0.8) +
  scale_x_continuous(breaks=seq(0,1,0.25),labels=c("0","0.25","0.5","0.75","1")) +
  scale_color_manual(values = Condition2.colours) +
  theme_bw() +
  theme(strip.text.x = element_text(size=10),
        legend.text.align = 0,
        legend.position = 'right', 
        legend.box.just = "left") +
  facet_wrap(~Lineage) +
  guides(color=guide_legend(title="Condition"))

p9 <- (pd1/pd2)
p9
```
# 5 Save figures
```{r}
celltype <- "RecM"
out <- here("data","plots",celltype)
if(!dir.exists(out)) {
  dir.create(out,recursive = TRUE)
}

showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)
# umap
layout = "AB\nAB"
p1 +
  p2 +
  plot_layout(design = layout) +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".umap.jpeg"),
       height=4,
       width=10)

# umap split by Condition
p3 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".umap.splitbyCondition.jpeg"),
       height=4,
       width=6.66)

# umap split by Condition2
p4 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,"TRM.umap.splitbyCondition2.jpeg"),
       height=3.5,
       width=10)


# cell proportion
p5 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".cellprop.jpeg"),
       height=4,
       width=8)

# dotplot
suppressWarnings(print(p6 +
  theme(plot.tag = element_text(size = 14, face = "bold",
                                family="Arial"
                                ))))
ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".dotplot.jpeg"),
       height=6,
       width=12)

suppressWarnings(print(p6 +
  theme(plot.tag = element_text(size = 14, face = "bold",
                                family="Arial"
                                ))))
# umap trajectory
p7 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".trajectory.jpeg"),
       height=4,
       width=8)

# Lineage 1 heatmap 
p8 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".geneExp_pseudotime.jpeg"),
       height=4,
       width=4)

# Pseudotime change
p9 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".pseudotimeChanges.jpeg"),
       height=5,
       width=10)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

```

# 6 Save object
```{r}
out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEU.qs"))

if (!file.exists(out)) {
  qsave(seu,file=out,nthreads=16)
}

```
