---
title: 'Analysis of the profibrotic RecM (MDM-PLA2G7) cluster'
author:
- name: Anson Wong
  affiliation: Molecular Immunity, Murdoch Children's Research Institute
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(dplyr)
library(ComplexHeatmap)
library(DOSE)
library(circlize)
library(tibble)
library(RColorBrewer)
library(ggplotify)
library(ensembldb)
library(Homo.sapiens)
library(msigdbr)
library(AnnotationHub)
library(scuttle)
library(fgsea)
library(dendsort)
library(readxl)
library(scattermore)
library(chroma)
library(cowplot)
library(ggrastr)
library(data.table)
library(gridExtra)
})
set.seed(1990)

```

# 2 Prepare reference object
## 2.1 Load object
```{r}
# load analysis object
celltype <- "RecM"
seu <- qread(here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEU.qs")),
             nthreads=16)
```

## 2.2 Prepare object
```{r}
# rename and extract metatdata
seu$orig.ident <- "CF paed"
metadata <- seu@meta.data[,c("orig.ident","nCount_RNA","nFeature_RNA","Age","Sex",
                     "batchID","subtype","Condition2","ID")]
colnames(metadata)
colnames(metadata) <- c("orig.ident","nCount_RNA","nFeature_RNA","age","sex",
                        "batchID","orig.celltype","condition","sample")

seu.ref <- CreateSeuratObject(seu@assays$RNA@counts, meta.data = metadata)

```

## 2.3 Set up colours
```{r}
subtype.colours <- setNames(c("#ef476f", "#7209b7", "#ffd166", "#83c5be",
                                 "#3a6ea5", "#9c6644", "#9e2a2b"),
                               levels(seu.ref$orig.celltype))
condition.colours <- setNames(c("#26547c","#ffd166","#ef476f"),
                               levels(seu.ref$condition))
```


# 3 Prepare public objects
## 3.1 Prepare ensembl v98 database
```{r}
ah <- AnnotationHub(cache="/group/canc2/anson/.cache/R/AnnotationHub", ask=FALSE)
EnsDb.Hsapiens.v98 <- query(ah, c("EnsDb", "Homo Sapiens", 98))[[1]]
```

## 3.2 Wendisch et al. dataset
```{r}
# load object
seu.wendisch <- qread(here("data","public_datasets","wendisch","SCEs","wendisch.SEU.qs"),nthreads=16)

wendisch.celltype <- c("Monocytes","Mono/Mφ","CD163/LGMN-Mφ")
# select columns from metadata
metadata <- seu.wendisch@meta.data %>%
  dplyr::mutate(orig.ident="Wendisch") %>%
  dplyr::mutate(batchID="Wendisch") %>%
  dplyr::mutate(condition="Covid") %>%
  dplyr::select(orig.ident, nCount_RNA, nFeature_RNA, age, sex,
                batchID, Cluster, condition, patient)
colnames(metadata)
colnames(metadata) <- c("orig.ident","nCount_RNA","nFeature_RNA","age","sex",
                        "batchID","orig.celltype","condition","sample")
metadata$orig.ident <- "Wendisch"

# match gene ID
ensdb_columns <- setNames(c("GENEBIOTYPE", "SEQNAME","SYMBOL"),
                          paste0("ENSEMBL.",
                                 c("GENEBIOTYPE", "SEQNAME","SYMBOL")))
stopifnot(all(ensdb_columns %in% columns(EnsDb.Hsapiens.v98)))

ensdb_df <- DataFrame(
  lapply(ensdb_columns, function(column) {
    mapIds(
      x = EnsDb.Hsapiens.v98,
      keys = rownames(seu.wendisch@assays$RNA@counts),
      keytype = "GENEID",
      column = column,
      multiVals = "first")
  }),
  row.names = rownames(seu.wendisch@assays$RNA@counts))

# make feature names unique
counts <- seu.wendisch@assays$RNA@counts
rownames(counts) <- uniquifyFeatureNames(rownames(ensdb_df), ensdb_df$ENSEMBL.SYMBOL)

# Create new object
seu.wendisch.backup <- seu.wendisch
seu.wendisch <- CreateSeuratObject(counts, meta.data = metadata)
seu.wendisch <- subset(seu.wendisch, orig.celltype %in% wendisch.celltype)

```

## 3.3 Li et al. dataset
```{r}
# Load objects
seu.li <- qread(here("data","public_datasets","li","SCEs","li.SEU.qs"),nthreads=16)
li.celltype <- c("Mono","FOLR2.IMs","SPP1.IMs")

# select columns from metadata
metadata <- seu.li@meta.data %>%
  dplyr::mutate(batchID="Li") %>%
  dplyr::mutate(orig.ident="Li") %>%
  dplyr::mutate(age="unknown") %>%
  dplyr::mutate(sex="unknown") %>%
  dplyr::mutate(sample=seu.li$orig.ident) %>%
  dplyr::mutate(condition=seu.li$orig.treatment) %>%
  dplyr::select(orig.ident, nCount_RNA, nFeature_RNA, age, sex,
                batchID, Cell.Types, orig.treatment, sample)
colnames(metadata)
colnames(metadata) <- c("orig.ident","nCount_RNA","nFeature_RNA","age","sex",
                        "batchID","orig.celltype","condition","sample")

# match gene ID
ensdb_columns <- setNames(c("GENEBIOTYPE", "SEQNAME","GENEID"),
                          paste0("ENSEMBL.",
                                 c("GENEBIOTYPE", "SEQNAME", "GENEID")))
stopifnot(all(ensdb_columns %in% columns(EnsDb.Hsapiens.v98)))

ensdb_df <- DataFrame(
  lapply(ensdb_columns, function(column) {
    mapIds(
      x = EnsDb.Hsapiens.v98,
      keys = rownames(seu.li@assays$RNA@counts),
      keytype = "SYMBOL",
      column = column,
      multiVals = "first")
  }),
  row.names = rownames(seu.li@assays$RNA@counts))

# make feature names unique
counts <- seu.li@assays$RNA@counts
rownames(counts) <- uniquifyFeatureNames(ensdb_df$ENSEMBL.GENEID, rownames(ensdb_df))

# Create new object
seu.li.backup <- seu.li
seu.li <- CreateSeuratObject(counts, meta.data = metadata)
seu.li <- subset(seu.li, orig.celltype %in% li.celltype)
seu.li$orig.celltype <- fct_drop(seu.li$orig.celltype)

```

# 4 Merge and save objects
```{r}
dim(seu.ref)
dim(seu.wendisch)
dim(seu.li)

# merge objects
seu <- merge(seu.ref, y=list(seu.wendisch,seu.li))
dim(seu)

# check number of cells per batch
seu$batchID <- factor(seu$batchID)
summary(seu$batchID)

if (!dir.exists(here("data","public_datasets","SCEs"))) {
  dir.create(here("data","public_datasets","SCEs"),recursive=TRUE)
}

out <- here("data","public_datasets","SCEs","public.processed.SEU.qs")
if (!file.exists(out)) {
  qsave(seu, file=out, nthreads=16)
}

```

# 5 Normalization, integration, clustering
For details, see code/bal_RecM.public_integration_clustering.R

# 6 Analysis
## 6.1 Gene set module score
Gene sets from 4 studies were selected:
- Li et al. 2022 [CF]
- Wendisch et al. 2021 [COVID-19]
- Liao et al. 2020 [COVID-19]
- Morrell et al. 2023 [AHRF]

### 6.1.1 Prepare gene sets
```{r}
# Li et al. 2022 [CF]
li.markers <- read_excel(here("data","public_datasets","li","inline-supplementary-material-2.xls"))
head(li.markers)

li.genesets <- 
  li.markers %>%
  dplyr::filter(cluster=="SPP1.IMs") %>%
  dplyr::select(gene) %>% unlist(use.names = FALSE)

# Wendisch et al. 2021 [Covid]
wendisch.genesets <- c("MARCKS","CD84","RAP2B","TMEM37","OLFML2B",
                       "MERTK","TTYH3","GPNMB","LGMN","RGL1",
                       "BLVRB","CTSB","NPL","TPCN1","CMKLR1",
                       "CD163","ABCA1","MAF","FPR3","PSAP",
                       "A2M","MPP1","GALM","LHFPL2","PLTP",
                       "BCAT1","SDC3","RNASE1","SLC36A1","FNIP2",
                       "PLXNC1","PMP22","RNF130","ETV5","MFSD1",
                       "MS4A6A","SDS","MAFB","CREG1","ADAM9",
                       "LIPA","BASP1","CPM","PLA2G7","AP2A2",
                       "ADAP2","CTSL","IL10","ADORA3","HM13")

# Liao et al. 2020 [Covid]
liao.markers <- read_excel(here("data","public_datasets","liao","Liao_tableS5.xlsx"))
head(liao.markers)

liao.genesets <-
  liao.markers %>%
  dplyr::filter(cluster=="Group3") %>%
  dplyr::select(gene) %>% unlist(use.names = FALSE)

# Morrell et al. 2023
morrell.markers <- read_excel(here("data","public_datasets","morrell","Morrell_tableS5.xlsx"))
head(morrell.markers)

morrell.genesets <- 
  morrell.markers %>% 
  data.frame() %>%
  dplyr::select(Cluster.0, Cluster.6) %>% unlist(use.names = FALSE)

gene.sets <- list("Li"=head(li.genesets,50),
                  "Wendisch"=wendisch.genesets,
                  "Liao"=head(liao.genesets,50),
                  "Morrell"=morrell.genesets)
```


### 6.1.2 Calculate gene set module score
```{r fig.width=12, fig.height=6}
seu.orig <- seu # read the original RecM object
seu.orig <- AddModuleScore(seu.orig, gene.sets)

min.score <- min(unlist(seu.orig@meta.data[,paste0("Cluster",1:4)],use.names = FALSE))
max.score <- max(unlist(seu.orig@meta.data[,paste0("Cluster",1:4)],use.names = FALSE))

subtype.colours <- setNames(c("#ef476f", "#7209b7", "#ffd166", "#83c5be","#3a6ea5", "#9c6644", "#9e2a2b"), # ref
                            levels(seu.orig$subtype))
subtype.labels <- c("Mono", 
                    "MDM-IV", 
                    expression('MDM-'*italic('CCR2')^'+'),
                    expression('MDM-'*italic('FABP4')^'low'),
                    expression('MDM-'*italic('FABP4')^'high'),
                    expression('MDM-'*italic('PLA2G7')^'high'),
                    "IM")

# feature plot of module score
p1 <- FeaturePlot(seu.orig, c(paste0("Cluster",1:4)),
            slot = "data",
            #keep.scale = "feature",
            ncol = 4,
            order = TRUE,
            raster=FALSE,
            pt.size =0.1) & 
  scale_colour_gradientn(colours = rev(brewer.rdylbu(11)),
                         limits = c(min.score, max.score), 
                         breaks = c(-0.5,0,0.5,1),
                         labels=c(-0.5,0,0.5,1)) &
  #NoLegend() &
  labs(x = NULL, y = NULL) &
  guides(x = "none", y = "none") &
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(size=10),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.1,"inch"),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

# violin plot of module score per subtype
title.list <- list(Cluster1=expression("Li"*italic('et al.')),
                   Cluster2=expression("Wendisch"*italic('et al.')),
                   Cluster3=expression("Liao"*italic('et al.')),
                   Cluster4=expression("Morrell"*italic('et al.')))
pt1.list <- list()
for (cluster in paste0("Cluster",1:4)) {
  p <- ggplot(seu.orig@meta.data, aes(x=subtype,y=!! sym(cluster))) +
        geom_jitter(mapping=aes(color = !! sym(cluster), alpha=0.5)) +
        geom_hline(yintercept = mean(unlist(seu.orig@meta.data[,cluster])),col = "#495057", size=1) +
        geom_violin(mapping=aes(fill=subtype)) +
        scale_fill_manual(values=subtype.colours) +
        scale_x_discrete(label=subtype.labels) +
        scale_colour_gradientn(colours = rev(brewer.rdylbu(11))) +
        theme_classic() +
        labs(title=title.list[[cluster]],x = NULL, y = "Score") +
        scale_y_continuous(limit=c(-0.5,1.2),breaks=c(-0.5,0,0.5,1),labels=c(-0.5,0,0.5,1)) +
        theme(plot.title = element_text(hjust = 0.5),
              #text=element_text(family="Arial"),
              axis.text.x = element_text(angle=45, vjust=1, hjust=1, size=10),
              #axis.ticks.x = element_blank(),
              axis.text.y= element_text(colour="black"),
              axis.title.y=element_text(vjust=-1,
                                        size=10),
              axis.title.x=element_blank(),
              legend.text = element_text(size = 10),
              strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
        NoLegend()
  pt1.list[[cluster]] <- p
}

p1 <- p1/(pt1.list[[1]] | pt1.list[[2]] | pt1.list[[3]] | pt1.list[[4]])
p1
```

### 6.1.3 One-sided wilcoxon test
```{r}
df <- split(seu.orig@meta.data, seu.orig$subtype)

# Li
stat.test1 <- seu.orig@meta.data %>%
  wilcox_test(Cluster1 ~ subtype,
              ref.group = "all",
              alternative="less",
              exact=TRUE,
              p.adjust.method = "BH",
              detailed = TRUE)

-log10(stat.test1$p.adj)

# Wendisch
stat.test2 <- seu.orig@meta.data %>%
  wilcox_test(Cluster2 ~ subtype,
              ref.group = "all",
              alternative="less",
              exact=TRUE,
              p.adjust.method = "BH",
              detailed = TRUE)

-log10(stat.test$p.adj)

# Liao
stat.test3 <- seu.orig@meta.data %>%
  wilcox_test(Cluster3 ~ subtype,
              ref.group = "all",
              alternative="less",
              exact=TRUE,
              p.adjust.method = "BH",
              detailed = TRUE)

-log10(stat.test$p.adj)

# Morrell
stat.test4 <- seu.orig@meta.data %>%
  wilcox_test(Cluster4 ~ subtype,
              ref.group = "all",
              alternative="less",
              exact=TRUE,
              p.adjust.method = "BH",
              detailed = TRUE)

-log10(stat.test$p.adj)

write.csv(stat.test1 %>% data.frame(), file=here("data","Li.stat.csv"))
write.csv(stat.test2 %>% data.frame(), file=here("data","Wendisch.stat.csv"))
write.csv(stat.test3 %>% data.frame(), file=here("data","Liao.stat.csv"))
write.csv(stat.test4 %>% data.frame(), file=here("data","Morrell.stat.csv"))

# write statistics
setDT(seu.orig@meta.data)
seu.orig@meta.data[, as.list(summary(dt)), by = subtype]

tapply(seu.orig@meta.data$Cluster1, seu.orig@meta.data$subtype,summary) %>% data.frame() %>% apply(2,as.character) %>% write.csv(file=here("data","Fig3A.stat.csv"))
```



## 6.2 Prepare clustered object
```{r}
seu <- qread(here("data","public_datasets","SCEs","public.integrated.clustered.SEU.qs"),nthreads=16)
seu@reductions$umap@cell.embeddings[,1] <- -seu@reductions$umap@cell.embeddings[,1]
seu@meta.data$orig.celltype[seu@meta.data$orig.ident == "Li" & seu@meta.data$orig.celltype == "Mono"] <- "Mono."

# rename celltype
seu$orig.celltype <- factor(seu$orig.celltype, levels=c("Mono","MDM-IV","MDM-CCR2+","MDM-FABP4low",
                                                        "MDM-FABP4high","MDM-PLA2G7high","IM",
                                                        "Mono.","SPP1.IMs","FOLR2.IMs",
                                                        "Monocytes","Mono/Mφ","CD163/LGMN-Mφ"))
# Split objects
## reference
seu.ref <- subset(seu, orig.ident == "CF paed")
seu.ref$orig.celltype <- fct_drop(seu.ref$orig.celltype)
seu.ref$orig.celltype <- factor(seu.ref$orig.celltype,
                                levels=c("Mono","MDM-IV","MDM-CCR2+","MDM-FABP4low",
                                         "MDM-FABP4high","MDM-PLA2G7high","IM"))
seu.ref$condition <- factor(seu.ref$condition, levels=c("HC","CFM","CFS"))

## Li et al.
seu.li <- subset(seu, orig.ident == "Li")
seu.li$orig.celltype <- fct_drop(seu.li$orig.celltype)
seu.li$orig.celltype <- factor(seu.li$orig.celltype,
                                levels=c("Mono.","SPP1.IMs","FOLR2.IMs"))

## Wendisch et al.
seu.wendisch <- subset(seu, orig.ident == "Wendisch")
seu.wendisch$orig.celltype <- fct_drop(seu.wendisch$orig.celltype)
seu.wendisch$orig.celltype <- factor(seu.wendisch$orig.celltype,
                                levels=c("Monocytes","Mono/Mφ","CD163/LGMN-Mφ"))

# set colors
dataset.colours <- setNames(c("#06948E","#C03830","#E7872B"),
                               levels(seu$orig.ident))
dataset.labels <- c("Present study", "Li", "Wendisch")

subtype.colours <- setNames(c("#ef476f", "#7209b7", "#ffd166", "#83c5be","#3a6ea5", "#9c6644", "#9e2a2b", # ref
                              "#d81159","#b08968","#99ccff", # Li et al.
                              "#ff002b","#f06900","#deab90" # Wendisch et al.
                              ),
                               levels(seu$orig.celltype))
```

## 6.3 UMAP embeddings
### 6.3.1 Grouped by datasets
```{r fig.height=3, fig.width=3}
# Grouped by datasets
p2 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.001,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        order = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=dataset.colours,
        #split.by = "Condition",
        group.by = "orig.ident") + #NoLegend() +
  labs(x = NULL, y = NULL,title="") + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.key.size = unit(0.2,"cm"),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) + NoLegend() +
  scale_color_manual(values = dataset.colours,
                    labels = dataset.labels)
suppressWarnings(print(p2))
```

### 6.3.2 Split by datasets; grouped by celltype labels
```{r fig.height=3, fig.width=12}
subtype.labels <- c("Mono", 
                    "MDM-IV", 
                    expression('MDM-'*italic('CCR2')^'+'),
                    expression('MDM-'*italic('FABP4')^'low'),
                    expression('MDM-'*italic('FABP4')^'high'),
                    expression('MDM-'*italic('PLA2G7')^'high'),
                    "IM",
                    expression(italic('SPP1')*".IMs"),
                    expression(italic('FOLR2')*".IMs"),
                    "Monocytes",
                    "Mono/Mφ",
                    expression(italic("CD163")*"/"*italic("LGMN")*"-Mφ"))

seu$orig.ident <- gsub("CF paed","Present study",seu$orig.ident)
seu$orig.ident <- factor(seu$orig.ident, levels=c("Present study","Li","Wendisch"))
dataset.colours <- setNames(c("#20a39e","#ef5b5b","#ffba49"),
                               levels(seu$orig.ident))

# ref
subtype.colours <- setNames(c("#ef476f", "#7209b7", "#ffd166", "#83c5be","#3a6ea5", "#9c6644", "#9e2a2b"), # ref
                            levels(seu.ref$orig.celltype))
subtype.labels <- c("Mono", 
                    "MDM-IV", 
                    expression('MDM-'*italic('CCR2')^'+'),
                    expression('MDM-'*italic('FABP4')^'low'),
                    expression('MDM-'*italic('FABP4')^'high'),
                    expression('MDM-'*italic('PLA2G7')^'high'),
                    "IM")

ps1 <- DimPlot(seu.ref, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        #split.by = "orig.ident",
        group.by = "orig.celltype") + #NoLegend() +
  labs(title="Present study", x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_text(size=12, hjust=0.5,face = "plain"),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
  scale_color_manual(values=subtype.colours, labels = subtype.labels)

# Li
subtype.colours <- setNames(c("#d81159", "#b08968","#a480cf"), # Li et al.
                            levels(seu.li$orig.celltype))
subtype.labels <- c("Mono", 
                    expression(italic('SPP1')*".IMs"),
                    expression(italic('FOLR2')*".IMs"))

ps2 <- DimPlot(seu.li, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        #split.by = "orig.ident",
        group.by = "orig.celltype") + #NoLegend() +
  labs(title="Li", x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_text(size=12, hjust=0.5, face="plain"),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
  scale_color_manual(values=subtype.colours, labels = subtype.labels)

# Wendisch
subtype.colours <- setNames(c("#ff002b","#218380","#deab90"), # Wendisch et al.
                            levels(seu.wendisch$orig.celltype))
subtype.labels <- c("Monocytes",
                    "Mono/Mφ",
                    expression(italic("CD163")*"/"*italic("LGMN")*"-Mφ"))

ps3 <- DimPlot(seu.wendisch, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        #split.by = "orig.ident",
        group.by = "orig.celltype") + #NoLegend() +
  labs(title="Wendisch", x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_text(size=12, hjust=0.5, face="plain"),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
  scale_color_manual(values=subtype.colours, labels = subtype.labels)

p3 <- suppressWarnings(print(ps1+ps2+ps3))
p3
```

### 6.3.3 Profibrotic cell type density
```{r fig.height=3, fig.width=9}
seu$Harmony.UMAP1 <- seu@reductions$umap@cell.embeddings[,1]
seu$Harmony.UMAP2 <- seu@reductions$umap@cell.embeddings[,2]
seu$CellID <- rownames(seu@meta.data)

get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

dens.umap.list <- list()
loop.cells <- levels(seu$orig.celltype)

for(i in seq_along(loop.cells)){
    i.c <- loop.cells[i]
    # change this to 1% of all cells
    i.n <- ceiling(sum(seu$orig.celltype %in% i.c)/100)
    if(i.n < 100){
      i.n <- 100
    }
    i.dens <- get_density(seu$Harmony.UMAP1[seu$orig.celltype %in% i.c],
                          seu$Harmony.UMAP2[seu$orig.celltype %in% i.c], n=i.n)
    
    dens.umap.list[[i.c]] <- data.frame("orig.celltype"=i.c,
                                        "CellID"=seu$CellID[seu$orig.celltype %in% i.c],
                                        "Dens"=i.dens)
}

umap.dens <- do.call(rbind.data.frame, dens.umap.list)
umap.dens.merge <- base::merge(seu@meta.data, umap.dens, by=c('orig.celltype', 'CellID'))


cell.order <- c("MDM-PLA2G7high","SPP1.IMs","CD163/LGMN-Mφ")
title.list <- setNames(c(expression('MDM-'*italic('PLA2G7')^'high'),
                         expression(italic('SPP1')*".IMs"),
                         expression(italic("CD163")*"/"*italic("LGMN")*"-Mφ")),
                       cell.order)
pt.list <- list()
for(x in seq_along(cell.order)){
  x.c <- cell.order[x]
  
  p <- ggplot(umap.dens.merge,
         aes(x=Harmony.UMAP1, y=Harmony.UMAP2)) +
    geom_scattermore(data=umap.dens.merge[, c("Harmony.UMAP1", "Harmony.UMAP2")],
                     colour='#adb5bd',) +
    geom_point_rast(data=umap.dens.merge[umap.dens.merge$orig.celltype %in% x.c, ],
                    aes(colour=Dens), size=0.001) +
    scale_colour_viridis() +
    labs(title=title.list[x.c], x="UMAP1", y="UMAP2") +
    theme_cowplot() +
    guides(colour=guide_colourbar(title="Density")) +
    theme(plot.title=element_text(size=8,hjust=0.5,face="plain"),
          legend.key.size = unit(0.2, "cm"),
          axis.text=element_blank(),
          axis.title=element_blank(),
          axis.line = element_blank(),
          axis.ticks=element_blank(),
          legend.title=element_text(size=6),
          legend.text=element_blank())
    pt.list[[x.c]] <- p
}


p4 <- pt.list[[1]] + pt.list[[2]] + pt.list[[3]]
p4
```

## 6.4 Feature violin plot
```{r}
subtype.labels <- c("Mono", 
                    "MDM-IV", 
                    expression('MDM-'*italic('CCR2')^'+'),
                    expression('MDM-'*italic('FABP4')^'low'),
                    expression('MDM-'*italic('FABP4')^'high'),
                    expression('MDM-'*italic('PLA2G7')^'high'),
                    "IM",
                    "Mono",
                    expression(italic('SPP1')*".IMs"),
                    expression(italic('FOLR2')*".IMs"),
                    "Monocytes",
                    "Mono/Mφ",
                    expression(italic("CD163")*"/"*italic("LGMN")*"-Mφ"))

Idents(seu) <- "orig.ident"

p <- VlnPlot(seu, c("SPP1","PLA2G7","CD163","LGMN","MARCKS","A2M","MERTK","GPNMB"),
        slot = "data",
        #split.plot = TRUE,
        combine=FALSE,
        group.by = "orig.celltype",
        #split.by = "orig.ident",
        #keep.scale = "feature",
        ncol = 4,
        pt.size =0)

p <- lapply(p, FUN = function(x) {
  pt <- x + 
    scale_fill_manual(values=subtype.colours) +
    scale_x_discrete(labels=subtype.labels) +
    labs(x = NULL) +
    theme(plot.title = element_text(face="bold.italic", size=12),
        axis.text.x=element_text(angle=45, vjust=1, hjust=1, size=8),
        axis.text.y=element_text(size=8)
        ) +
    NoLegend()
  return(pt)
})

p5 <- ((p[[1]] + theme(axis.text.x = element_blank(), 
                       axis.title.y=element_text(size=10,hjust=4))) |
  (p[[2]] + theme(axis.text.x = element_blank(), axis.title.y=element_blank())) |
  (p[[3]] + theme(axis.text.x = element_blank(), axis.title.y=element_blank())) |
  (p[[4]] + theme(axis.text.x = element_blank(), axis.title.y=element_blank())))/
  ((p[[5]] + theme(axis.title.y=element_blank())) | 
     (p[[6]] + theme(axis.title.y=element_blank())) |
     (p[[7]] + theme(axis.title.y=element_blank())) |
     (p[[8]] + theme(axis.title.y=element_blank())))

p5

```

# 7 Save figures
```{r}
celltype <- "public"
out <- here("data","public_datasets","plots")
if(!dir.exists(out)) {
  dir.create(out,recursive = TRUE)
}

showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

# gene set enrichment score
layout = "AAAAAAAA\nBBCCDDEE"
p1 +
  #plot_layout(design = layout) +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Arial"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","public.geneSetScore.jpeg"),
       height=6,
       width=12)

# umap grouped by dataset
p2 +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Arial"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","public.umap_dataset.jpeg"),
       height=4,
       width=4)

# umap split by dataset
layout = "AABBCC\nAABBCC"
ps1 + ps2 + ps3 +
  plot_layout(design = layout) +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Arial"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","public.umap_splitdataset.jpeg"),
       height=3,
       width=12)

# Kernel density plot
layout = "AABBCC"
pt.list[[1]]+pt.list[[2]]+pt.list[[3]] + 
  plot_layout(design = layout) +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Arial"))
ggsave(last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","public.kernelDensity.jpeg"),
       height=3,
       width=9)

# featureplot
p5 +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Arial"))
ggsave(last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","public.featurePlot.jpeg"),
       height=4,
       width=12)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

```


# 8 Save object
```{r}
out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEU.qs"))

if (!file.exists(out)) {
  qsave(seu,file=out,nthreads=16)
}

```



