---
title: 'Analysis of epithelial cells'
author:
- name: Anson Wong
  affiliation: Molecular Immunity, Murdoch Children's Research Institute
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(phateR)
library(slingshot)
library(tradeSeq)
library(org.Hs.eg.db)
library(ReactomePA)
library(clusterProfiler)
library(readr)
library(stringr)
})
set.seed(1990)

```

# 2 Analysis
## 2.1 Read processed object
```{r}
# load analysis object
celltype <- "Epithelial"
seu <- qread(here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs")),
             nthreads=16)
```

## 2.2 Prepare object for downstream analysis
```{r}
# Subset object
seu.backup <- seu
seu <- subset(seu.backup, Condition != "CF.ETI" & 
              sampleID != "M1C170C" & sampleID != "M1C176" & sampleID != "M1C180")
seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)
seu.backup <- NULL
gc()

# Add Condition2
draft1 <- factor(
      case_when(
        seu$Condition == "Control" ~ "HC"))
draft2 <- factor(
  case_when(
    seu$Bronchiectasis == "No" ~ "CFN",
    seu$Bronchiectasis == "Yes" ~ "CFB"
  )
)

Condition2 <- coalesce(draft1,draft2)
names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")

# Rename Condition
seu$Condition <- factor(
  case_when(
    seu$Condition == "Control" ~ "HC",
    seu$Condition == "CF" ~ "CF"
  ),
  levels=c("HC","CF")
)
# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", # healthy
                                "M1C170D","M1C176C","M1C177","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B", # CF mild
                                "M1C160(1)","M1C166(1)","M1C188","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A" # CF severe
                                ))

ID <- setNames(c(paste0("HC-",1:6),paste0("CF-N",1:8),paste0("CF-B",1:8)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],levels=c(paste0("HC-",1:6),paste0("CF-N",1:8),paste0("CF-B",1:8))),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Set colours
Condition2.colours <- setNames(c("#26547c","#ffd166","#ef476f"),
                               levels(seu$Condition2))
```

## 2.3 Differential gene expression analysis
### 2.3.1 Prepare object
Epithelial
```{r}
celltype <- "Epithelial"

# subset object
DefaultAssay(seu) <- "SCT"
s <- seu

```

Three comparisons:
CFN (n=8) vs HC (n=6)
```{r}
# set idents
Idents(s) <- "Condition2" # Change idents here

# define condition
test <- "CFN"
control <- "HC"

compare <- paste0(test,"_vs_",control)
if(!dir.exists(here("data", "DE", celltype, compare))) {
  dir.create(here("data", "DE", celltype, compare,"DEG"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"Reactome"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"WikiPathway"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_CC"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_BP"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_MF"), recursive = TRUE)
}
```

CFB (n=8) vs HC (n=6)
```{r}
# set idents
Idents(s) <- "Condition2" # Change idents here

# define condition
test <- "CFB"
control <- "HC"

compare <- paste0(test,"_vs_",control)
if(!dir.exists(here("data", "DE", celltype, compare))) {
  dir.create(here("data", "DE", celltype, compare,"DEG"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"Reactome"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"WikiPathway"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_CC"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_BP"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_MF"), recursive = TRUE)
}
```

CFB (n=8) vs CFN (n=6)
```{r}
# set idents
Idents(s) <- "Condition2" # Change idents here

# define condition
test <- "CFB"
control <- "CFN"

compare <- paste0(test,"_vs_",control)
if(!dir.exists(here("data", "DE", celltype, compare))) {
  dir.create(here("data", "DE", celltype, compare,"DEG"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"Reactome"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"WikiPathway"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_CC"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_BP"), recursive = TRUE)
  dir.create(here("data", "DE", celltype, compare,"GO_MF"), recursive = TRUE)
}
```

### 2.3.2 DE
DE results shown in Table E9 are written by code/writeDE.R
```{r}
# condition vs healthy
markers <- FindMarkers(s,
                       ident.1=test, # condition
                       ident.2=control, # healthy
                       test.use="MAST",
                       random.seed=1990,
                       recorrect_umi=FALSE,
                       min.pct=0.1,
                       logfc.threshold = 0.25,
                       only.pos = FALSE)

markers$gene <- rownames(markers)

# get sig up- and down-regulated genes
deg <- markers %>% dplyr::filter(p_val_adj < 0.05)

up <- deg %>% dplyr::filter(p_val_adj < 0.05 & avg_log2FC > 0.25) %>% dplyr::arrange(desc(avg_log2FC))
down <- deg %>% dplyr::filter(p_val_adj < 0.05 & avg_log2FC < -0.25) %>% dplyr::arrange(avg_log2FC)

# write results
write.csv(up %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-up.csv")),
            row.names = FALSE)
write.csv(down %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-down.csv")),
            row.names = FALSE)
write.csv(deg %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-all.csv")),
            row.names = FALSE)
write.csv(markers %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-raw.csv")),
            row.names = FALSE)
```

### 2.3.3 GO enrichment
Top 10 results shown in Table E11 are written by code/bal_GOenrichment.R
```{r}
# get ensemblDB
getCols <- setNames(c("SYMBOL","ENTREZID"),c("SYMBOL","ENTREZID"))

genes <- data.frame(
  lapply(getCols, function(column) {
    mapIds(
      x = org.Hs.eg.db,
      keys = unique(deg$gene),
      keytype = "SYMBOL",
      column = column)
  }),
  row.names = unique(deg$gene))

# GO BP enrichment
go.up.bp <- enrichGO(genes[up$gene,]$ENTREZID, OrgDb=org.Hs.eg.db, ont="BP",
                       pAdjustMethod = "BH", readable=TRUE, pvalueCutoff = 0.01)

go.down.bp <- enrichGO(genes[down$gene,]$ENTREZID, OrgDb=org.Hs.eg.db, ont="BP",
                     pAdjustMethod = "BH", readable=TRUE, pvalueCutoff = 0.01)

# Remove redundant GO terms
go.up.bp <- clusterProfiler::simplify(go.up.bp)
go.down.bp <- clusterProfiler::simplify(go.down.bp)

# write results
go.up.bp@result %>%
  dplyr::rename(Pathway=Description) %>%
  dplyr::mutate(Celltype = celltype) %>%
  dplyr::mutate(Comparison=compare) %>%
  dplyr::mutate(Direction="Up") %>%
  dplyr::mutate(Genes=gsub("/","; ",go.up.bp@result$geneID)) %>%
  dplyr::select(Pathway, Celltype, Comparison, Direction, ID, Count, pvalue, p.adjust, Genes) %>%
  slice_head(n = 10) -> up.df
  write_csv(up.df, file = here("data","GOBP",celltype,compare, "GO_BP",
                        paste0(celltype,".",compare,".GO_BP-up.csv")))

go.down.bp@result %>%
  dplyr::rename(Pathway=Description) %>%
  dplyr::mutate(Celltype = celltype) %>%
  dplyr::mutate(Comparison=compare) %>%
  dplyr::mutate(Direction="Down") %>%
  dplyr::mutate(Genes=gsub("/","; ",go.down.bp@result$geneID)) %>%
  dplyr::select(Pathway, Celltype, Comparison, Direction, ID, Count, pvalue, p.adjust, Genes) %>%
  slice_head(n = 10) -> down.df
  write_csv(down.df, file = here("data","GOBP",celltype,compare, "GO_BP",
                        paste0(celltype,".",compare,".GO_BP-down.csv")))
```

### 2.3.4 Violin plots of selected DEG
```{r fig.width=9, fig.height=3}
cilium <- c("SPAG16","DNAH11","RFX3","CFAP157","DNAH3","DNAH7","CFAP54","CFAP43","SPEF2","EFHC1","CFAP69","CFAP52","ARMC2","DNAI1","DNAH5","BBS4","DNAH6")
antigen <- c("B2M", "HLA-A", "HLA-B","HLA-DRA","HLA-DRB1","CTSL","CTSC","CD55","CD74","PSMB8","PSME1","CACLR","SCGB1A1")
genes <- c(cilium, antigen)

p1 <- DotPlot(seu,
            features = genes,
            group.by = "Condition2",
            col.max = 1,
            col.min = -1,
            dot.scale = 6
            ) + #coord_flip() +
    geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
    FontSize(y.text = 10, x.text = 8) + 
    labs(y = element_blank(), x = element_blank()) + 
      theme_bw() +
    theme(
          axis.text.x = element_text(size=10,
                                     face="italic",
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text.y = element_text(size=10,hjust=0),
          legend.direction="horizontal",
          legend.position="right",
          legend.justification = "right",
          legend.box="vertical",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()
          ) +
      
    scale_color_gradientn(colours = rev(brewer.rdbu(11)[2:10])) +
    #scale_y_discrete(labels = rev(c("CF.ETI.BL","CF.ETI","HC"))) +
    scale_size_continuous(limits = c(0, 100),
                          breaks = c(0, 25, 50, 75, 100),
                          labels=c(0,"",50,"",100)) +
    guides(size=guide_legend(override.aes=list(shape=21, 
                                               colour="black", 
                                               fill="#adb5bd"),
                                               title="% of exp. cells",
                                               label.position="bottom",
                                               title.position="top",
                                               title.hjust=0.5
                                               ))
p1$guides$colour$title <- "Mean Expression"
p1$guides$colour$title.position <- "top"
p1$guides$colour$title.hjust <- 0.5

suppressWarnings(p1)

```

## 2.4 Pathway activity scoring
```{r}
#### 4.6.1.3 Pathway activity score
lappend <- function (lst, ...){
lst <- c(lst, list(...))
  return(lst)
}

c5_sets <- read.gmt('/group/canc2/anson/reference/annotation/gsea-msigdb/msigdb_v2023.1.Hs_GMTs/c5.all.v2023.1.Hs.symbols.gmt')

# read GO files
celltype <- "Epithelial"
test <- "CFB"
control <- "HC"
compare <- paste0(test,"_vs_",control)
go.up.bp <- read.csv(here("data","GOBP",celltype,compare, "GO_BP",
                          paste0(celltype,".",compare,".GO_BP-up.csv")),
                     header = TRUE)
go.up.down <- read.csv(here("data","GOBP",celltype,compare, "GO_BP",
                          paste0(celltype,".",compare,".GO_BP-down.csv")),
                     header = TRUE)

# extract union of DEGs from the top 10 enriched pathways
set1 <- Reduce(union,sapply(go.up.bp[1:10,"Genes"],FUN=function(x) {str_split(x,"; ")}))
set2 <- Reduce(union,sapply(go.up.down[1:10,"Genes"],FUN=function(x) {str_split(x,"; ")}))

sets <- list(set1, set2)

DefaultAssay(seu) <- "SCT"
seu <- AddModuleScore(seu, features=sets, assay = "SCT", name="Epi")

seu$`Cilium activity` <- seu$Epi1
seu$`Barrier immunity` <- seu$Epi2

seu$Epi1 <- NULL
seu$Epi2 <- NULL

# t-test
df <- seu@meta.data[,c("Cilium activity","Barrier immunity","Condition2")] %>% 
  data.frame() %>%
  tidyr::pivot_longer(
    cols=c("Cilium.activity","Barrier.immunity"))
colnames(df) <- c("Condition2","Pathway","Score")
df$Pathway <- factor(gsub("\\."," ",df$Pathway),levels=c("Cilium activity",
                                                         "Barrier immunity"))

stat.test <- df %>%
  group_by(Pathway) %>%
  wilcox_test(Score ~ Condition2) %>%
  adjust_pvalue(method="BH") %>%
  add_significance()

stat.test <- stat.test %>%
  add_xy_position(x = "Condition2", dodge = 0.8)

stat.test$p.adj <- signif(stat.test$p.adj,digits = 3)

# box plot
p2 <- ggviolin(df, x="Condition2",
         y="Score", facet.by = "Pathway",
         fill="Condition2",
         palette = c("#26547c", "#ffd166", "#ef476f"),
         repel=TRUE) +
  stat_pvalue_manual(stat.test, 
                     label = "p.adj.signif", 
                     tip.length = 0.02,
                     label.size = 2.5,
                     #y.position = 2.5, 
                     step.increase = 0.02,
                     bracket.nudge.y = 0.4, 
                     vjust = 0, 
                     hide.ns = TRUE,
                     ) +
  labs(y="Pathway score") + NoLegend() +
  scale_x_discrete(labels=c("HC","CFN","CFB")) +
  theme(axis.text.x = element_text(#angle = 45,
                                   #vjust = 1,
                                   #hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.title.x=element_blank(),
        strip.background =element_rect(fill="#e5e5e5"),
        strip.text =element_text(size=14,face="bold"),
        axis.text.y = element_text(size=10,
                                   colour = "black"))
p2
```


# 3 Save figures
```{r}
celltype <- "Epithelial"
out <- here("data","plots",celltype)
if(!dir.exists(out)) {
  dir.create(out,recursive = TRUE)
}
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

# dotplot of DEG
p1 +
  theme(plot.tag = element_text(size = 14, face = "bold"))
ggsave(p1,
       filename=paste0(out,"/",paste0("bal_",celltype,".dotplot.jpeg")),
       device="jpeg",
       dpi=1600,
       height=2.5,
       width=10)

# pathway module score
p2 +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_Epithelial.pathwayActivity.jpeg"),
       height=4,
       width=6)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)
```

# 4 Save object
```{r}
out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.",date,".SEU.qs"))

if (!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

```
